---
title: "NCBI vs Ensembl Gene Names & Human Orthologs"
author: "Generated by Claude Code"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# 1. Setup & Data Loading

```{r load_libraries}
library(tidyverse)
library(scales)
library(patchwork)

theme_set(theme_minimal(base_size = 12))

# Classification color palette (reused throughout)
class_colors <- c(
  "Green" = "#2ecc71", "Yellow" = "#f39c12",
  "Red" = "#e74c3c", "NotMapped" = "#95a5a6"
)
```

```{r load_metadata}
# Gene metadata from preprocessing script
data_dir <- "../experiments/gene_name_analysis"
ens_meta <- read_tsv(file.path(data_dir, "ens_gene_metadata.tsv"), show_col_types = FALSE)
ncbi_meta <- read_tsv(file.path(data_dir, "ncbi_gene_metadata.tsv"), show_col_types = FALSE)

# Define "named" flags
ens_meta <- ens_meta %>%
  mutate(is_named = name != "NA" & !is.na(name))

ncbi_meta <- ncbi_meta %>%
  mutate(is_named = !is_loc & name != "NA" & !is.na(name))
```

```{r load_comparison}
# Comparison data (gene rows only)
comparison_file <- "../experiments/comparison_runs/ncbi_vs_ensembl_Ssal_v3.1/comparison_v3.tsv"
comparison <- read_tsv(comparison_file, show_col_types = FALSE)

genes_comp <- comparison %>%
  filter(feature == "gene") %>%
  mutate(
    class = str_extract(stats, "class=([^;]+)") %>% str_remove("class="),
    class_initial = str_extract(stats, "class_initial=([^;]+)") %>% str_remove("class_initial="),
    best_jaccard_exon = str_extract(stats, "best_jaccard_exon=([0-9.]+)") %>%
      str_remove("best_jaccard_exon=") %>% as.numeric(),
    best_jaccard_cds = str_extract(stats, "best_jaccard_cds_phase=([0-9.]+)") %>%
      str_remove("best_jaccard_cds_phase=") %>% as.numeric(),
    partner_countA = str_extract(stats, "partner_countA=([0-9]+)") %>%
      str_remove("partner_countA=") %>% as.integer(),
    partner_countB = str_extract(stats, "partner_countB=([0-9]+)") %>%
      str_remove("partner_countB=") %>% as.integer(),
    overlap_bp = str_extract(stats, "overlap_bp=([0-9]+)") %>%
      str_remove("overlap_bp=") %>% as.integer()
  )

# Strip "gene:" and "gene-" prefixes to match metadata gene_id format
genes_comp <- genes_comp %>%
  mutate(
    ncbi_id = geneA,
    ens_id = str_remove(geneB, "^gene:")
  )
```

```{r load_orthologs}
# Human orthologs
ens_ortho <- read_tsv(file.path(data_dir, "ens_human_orthologs.tsv"), show_col_types = FALSE)
ncbi_ortho <- read_tsv(file.path(data_dir, "ncbi_human_orthologs.tsv"), show_col_types = FALSE)

# Summarise to one row per salmon gene (may have multiple human orthologs)
ens_ortho_summary <- ens_ortho %>%
  group_by(ensembl_gene_id) %>%
  summarise(
    n_human_orthologs = n(),
    human_genes = paste(unique(hsapiens_homolog_associated_gene_name), collapse = ","),
    orthology_types = paste(unique(hsapiens_homolog_orthology_type), collapse = ","),
    .groups = "drop"
  )

ncbi_ortho_summary <- ncbi_ortho %>%
  group_by(salmon_gene_id) %>%
  summarise(
    n_human_orthologs = n(),
    human_genes = paste(unique(human_symbol), collapse = ","),
    .groups = "drop"
  )
```

```{r join_data}
# Join comparison data with metadata from both annotations
genes_joined <- genes_comp %>%
  left_join(
    ens_meta %>% select(gene_id, ens_biotype = biotype, ens_name = name,
                        ens_named = is_named, ens_description = description,
                        ens_desc_source = description_source,
                        ens_feature_type = feature_type),
    by = c("ens_id" = "gene_id")
  ) %>%
  left_join(
    ncbi_meta %>% select(gene_id, ncbi_biotype = biotype, ncbi_name = name,
                         ncbi_named = is_named, ncbi_is_loc = is_loc,
                         ncbi_gene_id_num = ncbi_gene_id,
                         ncbi_description = description),
    by = c("ncbi_id" = "gene_id")
  )

# Add ortholog info
genes_joined <- genes_joined %>%
  left_join(
    ens_ortho_summary %>% select(ensembl_gene_id, ens_has_ortholog = n_human_orthologs,
                                  ens_human_genes = human_genes),
    by = c("ens_id" = "ensembl_gene_id")
  ) %>%
  left_join(
    ncbi_ortho_summary %>% select(salmon_gene_id, ncbi_has_ortholog = n_human_orthologs,
                                   ncbi_human_genes = human_genes),
    by = c("ncbi_gene_id_num" = "salmon_gene_id")
  ) %>%
  mutate(
    ens_has_ortholog = !is.na(ens_has_ortholog),
    ncbi_has_ortholog = !is.na(ncbi_has_ortholog)
  )
```

# 2. Gene Annotation Overview

## Total gene counts by annotation and biotype

```{r overview_counts}
# Ensembl biotype summary
ens_biotype <- ens_meta %>%
  count(biotype, is_named) %>%
  mutate(annotation = "Ensembl")

ncbi_biotype <- ncbi_meta %>%
  count(biotype, is_named) %>%
  mutate(annotation = "NCBI")

biotype_combined <- bind_rows(ens_biotype, ncbi_biotype)

# Summary table
biotype_summary <- biotype_combined %>%
  group_by(annotation, biotype) %>%
  summarise(total = sum(n), named = sum(n[is_named]), .groups = "drop") %>%
  mutate(pct_named = named / total * 100) %>%
  arrange(annotation, desc(total))

biotype_summary %>%
  knitr::kable(
    caption = "Gene counts by biotype and annotation",
    col.names = c("Annotation", "Biotype", "Total", "Named", "% Named"),
    digits = 1
  )
```

```{r overview_stacked_bar, fig.height=7}
# Top biotypes for readability
top_biotypes <- biotype_summary %>%
  group_by(biotype) %>%
  summarise(total = sum(total)) %>%
  arrange(desc(total)) %>%
  head(10) %>%
  pull(biotype)

biotype_combined %>%
  filter(biotype %in% top_biotypes) %>%
  mutate(
    name_status = if_else(is_named, "Named", "Unnamed"),
    biotype = factor(biotype, levels = rev(top_biotypes))
  ) %>%
  ggplot(aes(x = biotype, y = n, fill = name_status)) +
  geom_col(position = "stack") +
  facet_wrap(~annotation) +
  coord_flip() +
  scale_fill_manual(values = c("Named" = "#3498db", "Unnamed" = "#bdc3c7")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Gene Counts by Biotype: Named vs Unnamed",
    x = "Biotype",
    y = "Number of genes",
    fill = ""
  )
```

## Key numbers

```{r key_numbers}
ens_total <- nrow(ens_meta)
ens_named_total <- sum(ens_meta$is_named)
ncbi_total <- nrow(ncbi_meta)
ncbi_named_total <- sum(ncbi_meta$is_named)

ens_pc_named <- sum(ens_meta$is_named & ens_meta$biotype == "protein_coding")
ncbi_pc_named <- sum(ncbi_meta$is_named & ncbi_meta$biotype == "protein_coding")

cat(sprintf("Ensembl: %s total genes, %s named (%.1f%%)\n",
            comma(ens_total), comma(ens_named_total), ens_named_total/ens_total*100))
cat(sprintf("  Protein-coding named: %s\n", comma(ens_pc_named)))
cat(sprintf("NCBI: %s total genes, %s with real symbols (%.1f%%)\n",
            comma(ncbi_total), comma(ncbi_named_total), ncbi_named_total/ncbi_total*100))
cat(sprintf("  Protein-coding named: %s\n", comma(ncbi_pc_named)))
cat(sprintf("\nEnsembl lncRNA genes: %s (essentially never named)\n",
            comma(sum(ens_meta$biotype == "lncRNA"))))
```


# 3. Gene Name Coverage Comparison

```{r naming_crosstab}
# Focus on matched genes (those in comparison data)
matched <- genes_joined %>%
  filter(class %in% c("Green", "Yellow", "Red"))

# Name status cross-tabulation
name_cross <- matched %>%
  mutate(
    ens_status = if_else(ens_named, "Ensembl named", "Ensembl unnamed"),
    ncbi_status = if_else(ncbi_named, "NCBI named", "NCBI unnamed")
  ) %>%
  count(ens_status, ncbi_status, class)

# Summary table
name_summary <- matched %>%
  mutate(
    name_category = case_when(
      ens_named & ncbi_named ~ "Both named",
      ens_named & !ncbi_named ~ "Ensembl only",
      !ens_named & ncbi_named ~ "NCBI only",
      TRUE ~ "Neither named"
    )
  ) %>%
  count(name_category, class) %>%
  pivot_wider(names_from = class, values_from = n, values_fill = 0) %>%
  mutate(Total = Green + Yellow + Red)

name_summary %>%
  knitr::kable(caption = "Name status of matched gene pairs by classification")
```

```{r naming_heatmap, fig.width=8, fig.height=6}
# Cross-tabulation heatmap by classification
name_cross_wide <- matched %>%
  mutate(
    name_category = case_when(
      ens_named & ncbi_named ~ "Both named",
      ens_named & !ncbi_named ~ "Ensembl only",
      !ens_named & ncbi_named ~ "NCBI only",
      TRUE ~ "Neither"
    ),
    class = factor(class, levels = c("Green", "Yellow", "Red"))
  ) %>%
  count(name_category, class) %>%
  mutate(name_category = factor(name_category,
                                levels = c("Both named", "Ensembl only", "NCBI only", "Neither")))

ggplot(name_cross_wide, aes(x = class, y = name_category, fill = n)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = comma(n)), size = 5) +
  scale_fill_gradient(low = "#f0f0f0", high = "#3498db", labels = comma) +
  labs(
    title = "Gene Pair Name Status by Classification",
    x = "Classification",
    y = "Name status",
    fill = "Count"
  )
```

```{r naming_notmapped}
# NotMapped genes: naming rate comparison
notmapped <- genes_joined %>% filter(class == "NotMapped")
mapped <- genes_joined %>% filter(class != "NotMapped")

naming_by_mapping <- tibble(
  Category = c("Mapped (Green/Yellow/Red)", "NotMapped"),
  Ens_named = c(
    mean(mapped$ens_named, na.rm = TRUE),
    mean(notmapped$ens_named, na.rm = TRUE)
  ),
  NCBI_named = c(
    mean(mapped$ncbi_named, na.rm = TRUE),
    mean(notmapped$ncbi_named, na.rm = TRUE)
  )
)

naming_by_mapping %>%
  pivot_longer(-Category, names_to = "Annotation", values_to = "Fraction_named") %>%
  mutate(Annotation = str_replace(Annotation, "_", " ")) %>%
  ggplot(aes(x = Category, y = Fraction_named, fill = Annotation)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  scale_fill_manual(values = c("Ens named" = "#9b59b6", "NCBI named" = "#3498db")) +
  labs(
    title = "Fraction of Named Genes: Mapped vs NotMapped",
    x = "",
    y = "Fraction with real gene name",
    fill = ""
  )
```


# 4. Why Are Genes Unnamed?

## Named in Ensembl, unnamed in NCBI

```{r ens_named_ncbi_unnamed}
ens_only <- matched %>%
  filter(ens_named & !ncbi_named)

cat(sprintf("Genes named in Ensembl but not NCBI: %s\n", comma(nrow(ens_only))))

# Biotype distribution
ens_only %>%
  count(ens_biotype, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100) %>%
  head(10) %>%
  knitr::kable(
    caption = "Biotype distribution of Ensembl-named, NCBI-unnamed genes",
    col.names = c("Ensembl biotype", "Count", "%"),
    digits = 1
  )
```

```{r ens_only_classification}
# Classification quality
ens_only %>%
  count(class) %>%
  mutate(pct = n / sum(n) * 100) %>%
  knitr::kable(
    caption = "Classification of Ensembl-named, NCBI-unnamed genes",
    col.names = c("Class", "Count", "%"),
    digits = 1
  )
```

```{r ens_only_ncbi_description}
# Do NCBI LOC partners have matching descriptions?
ens_only_desc <- ens_only %>%
  filter(class == "Green") %>%
  mutate(
    ncbi_has_desc = ncbi_description != "NA" & !is.na(ncbi_description) &
                    !str_detect(replace_na(ncbi_description, ""), "^uncharacterized LOC")
  )

cat(sprintf("Green-class, Ensembl-named, NCBI LOC genes: %d\n", nrow(ens_only_desc)))
cat(sprintf("  NCBI partner has non-trivial description: %d (%.1f%%)\n",
            sum(ens_only_desc$ncbi_has_desc, na.rm = TRUE),
            mean(ens_only_desc$ncbi_has_desc, na.rm = TRUE) * 100))
```

```{r ens_only_desc_source}
# Description source analysis
ens_only %>%
  count(ens_desc_source, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100) %>%
  knitr::kable(
    caption = "Ensembl description source for genes named only in Ensembl",
    col.names = c("Source", "Count", "%"),
    digits = 1
  )
```

## Named in NCBI, unnamed in Ensembl

```{r ncbi_named_ens_unnamed}
ncbi_only <- matched %>%
  filter(!ens_named & ncbi_named)

cat(sprintf("Genes named in NCBI but not Ensembl: %s\n", comma(nrow(ncbi_only))))

# Ensembl biotype breakdown
ncbi_only %>%
  count(ens_biotype, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100) %>%
  head(10) %>%
  knitr::kable(
    caption = "Ensembl biotype of NCBI-named, Ensembl-unnamed genes",
    col.names = c("Ensembl biotype", "Count", "%"),
    digits = 1
  )
```

```{r ncbi_only_classification}
ncbi_only %>%
  count(class) %>%
  mutate(pct = n / sum(n) * 100) %>%
  knitr::kable(
    caption = "Classification of NCBI-named, Ensembl-unnamed genes",
    col.names = c("Class", "Count", "%"),
    digits = 1
  )
```

```{r ncbi_only_ens_desc}
# Does Ensembl have description but no Name?
ncbi_only_ens_desc <- ncbi_only %>%
  mutate(
    ens_has_desc = ens_description != "NA" & !is.na(ens_description)
  )

cat(sprintf("NCBI-named, Ensembl-unnamed genes: %d\n", nrow(ncbi_only)))
cat(sprintf("  Ensembl partner has description but no Name: %d (%.1f%%)\n",
            sum(ncbi_only_ens_desc$ens_has_desc),
            mean(ncbi_only_ens_desc$ens_has_desc) * 100))
```

## Genes unnamed in both

```{r unnamed_both}
neither_named <- matched %>%
  filter(!ens_named & !ncbi_named)

cat(sprintf("Genes unnamed in both annotations: %s\n", comma(nrow(neither_named))))

# Biotype and classification
neither_named %>%
  count(ens_biotype, class) %>%
  pivot_wider(names_from = class, values_from = n, values_fill = 0) %>%
  mutate(Total = Green + Yellow + Red) %>%
  arrange(desc(Total)) %>%
  head(10) %>%
  knitr::kable(caption = "Unnamed in both: biotype x classification")
```


# 5. Gene Name Concordance

```{r name_concordance}
both_named <- matched %>%
  filter(ens_named & ncbi_named)

cat(sprintf("Gene pairs where both are named: %s\n", comma(nrow(both_named))))

both_named <- both_named %>%
  mutate(
    names_match_exact = ens_name == ncbi_name,
    names_match_ci = str_to_lower(ens_name) == str_to_lower(ncbi_name),
    match_type = case_when(
      names_match_exact ~ "Exact match",
      names_match_ci ~ "Case-insensitive match",
      TRUE ~ "Different"
    )
  )

concordance <- both_named %>%
  count(match_type) %>%
  mutate(pct = n / sum(n) * 100)

concordance %>%
  knitr::kable(
    caption = "Gene name concordance among pairs where both are named",
    col.names = c("Match type", "Count", "%"),
    digits = 1
  )
```

```{r concordance_by_class}
both_named %>%
  count(match_type, class) %>%
  mutate(class = factor(class, levels = c("Green", "Yellow", "Red"))) %>%
  ggplot(aes(x = class, y = n, fill = match_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("Exact match" = "#2ecc71",
                                "Case-insensitive match" = "#f39c12",
                                "Different" = "#e74c3c")) +
  labs(
    title = "Gene Name Concordance by Classification",
    x = "Classification",
    y = "Proportion",
    fill = "Name match"
  )
```

```{r discordant_examples}
# Examples of interesting discordances
discordant <- both_named %>%
  filter(match_type == "Different") %>%
  select(ncbi_id, ens_id, ncbi_name, ens_name, class,
         best_jaccard_exon, best_jaccard_cds) %>%
  arrange(desc(best_jaccard_cds))

cat(sprintf("Total discordant name pairs: %d\n\n", nrow(discordant)))

discordant %>%
  head(20) %>%
  knitr::kable(
    caption = "Top discordant gene names (highest CDS Jaccard)",
    col.names = c("NCBI ID", "Ensembl ID", "NCBI name", "Ensembl name",
                  "Class", "Jaccard exon", "Jaccard CDS"),
    digits = 3
  )
```


# 6. Human Ortholog Analysis

## Ortholog coverage

```{r ortholog_coverage}
# How many salmon genes have human orthologs by source
n_ens_with_ortho <- n_distinct(ens_ortho$ensembl_gene_id)
n_ncbi_with_ortho <- n_distinct(ncbi_ortho$salmon_gene_id)

cat(sprintf("Ensembl genes with human orthologs: %s (%.1f%% of %s)\n",
            comma(n_ens_with_ortho),
            n_ens_with_ortho / ens_total * 100,
            comma(ens_total)))
cat(sprintf("NCBI genes with human orthologs: %s (%.1f%% of %s)\n",
            comma(n_ncbi_with_ortho),
            n_ncbi_with_ortho / ncbi_total * 100,
            comma(ncbi_total)))
```

## Ortholog vs naming correlation

```{r ortholog_naming, fig.width=10, fig.height=5}
# For Ensembl: ortholog vs named
ens_ortho_naming <- ens_meta %>%
  mutate(has_ortholog = gene_id %in% ens_ortho_summary$ensembl_gene_id) %>%
  count(is_named, has_ortholog) %>%
  mutate(annotation = "Ensembl")

# For NCBI: ortholog vs named
ncbi_ortho_naming <- ncbi_meta %>%
  mutate(
    ncbi_gene_id_num = as.numeric(ncbi_gene_id),
    has_ortholog = ncbi_gene_id_num %in% ncbi_ortho_summary$salmon_gene_id
  ) %>%
  count(is_named, has_ortholog) %>%
  mutate(annotation = "NCBI")

ortho_naming <- bind_rows(ens_ortho_naming, ncbi_ortho_naming) %>%
  mutate(
    label = case_when(
      is_named & has_ortholog ~ "Named + ortholog",
      is_named & !has_ortholog ~ "Named, no ortholog",
      !is_named & has_ortholog ~ "Unnamed + ortholog",
      TRUE ~ "Unnamed, no ortholog"
    ),
    label = factor(label, levels = c("Named + ortholog", "Named, no ortholog",
                                      "Unnamed + ortholog", "Unnamed, no ortholog"))
  )

ggplot(ortho_naming, aes(x = annotation, y = n, fill = label)) +
  geom_col(position = "stack") +
  geom_text(aes(label = comma(n)), position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = c(
    "Named + ortholog" = "#2ecc71",
    "Named, no ortholog" = "#85c1e9",
    "Unnamed + ortholog" = "#f39c12",
    "Unnamed, no ortholog" = "#bdc3c7"
  )) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Gene Naming vs Human Ortholog Status",
    x = "",
    y = "Number of genes",
    fill = ""
  )
```

```{r ortholog_odds_ratio}
# Odds ratio: having ortholog and being named
# Ensembl
ens_or_data <- ens_meta %>%
  mutate(has_ortholog = gene_id %in% ens_ortho_summary$ensembl_gene_id)
ens_tbl <- table(named = ens_or_data$is_named, ortholog = ens_or_data$has_ortholog)
ens_or <- fisher.test(ens_tbl)

# NCBI
ncbi_or_data <- ncbi_meta %>%
  mutate(has_ortholog = as.numeric(ncbi_gene_id) %in% ncbi_ortho_summary$salmon_gene_id)
ncbi_tbl <- table(named = ncbi_or_data$is_named, ortholog = ncbi_or_data$has_ortholog)
ncbi_or <- fisher.test(ncbi_tbl)

cat(sprintf("Ensembl: OR = %.1f (95%% CI: %.1f-%.1f), p < %.1e\n",
            ens_or$estimate, ens_or$conf.int[1], ens_or$conf.int[2], ens_or$p.value))
cat(sprintf("NCBI:    OR = %.1f (95%% CI: %.1f-%.1f), p < %.1e\n",
            ncbi_or$estimate, ncbi_or$conf.int[1], ncbi_or$conf.int[2], ncbi_or$p.value))
```

## Unnamed genes with human orthologs (naming candidates)

```{r unnamed_with_orthologs}
# Ensembl unnamed genes that have human orthologs
ens_unnamed_ortho <- ens_meta %>%
  filter(!is_named) %>%
  inner_join(ens_ortho_summary, by = c("gene_id" = "ensembl_gene_id"))

cat(sprintf("Ensembl unnamed genes with human ortholog: %d\n", nrow(ens_unnamed_ortho)))

ens_unnamed_ortho %>%
  count(biotype, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100) %>%
  head(10) %>%
  knitr::kable(
    caption = "Biotypes of unnamed Ensembl genes with human orthologs",
    col.names = c("Biotype", "Count", "%"),
    digits = 1
  )
```

```{r ncbi_unnamed_orthologs}
# NCBI unnamed (LOC) genes with human orthologs
ncbi_unnamed_ortho <- ncbi_meta %>%
  filter(!is_named) %>%
  mutate(ncbi_gene_id_num = as.numeric(ncbi_gene_id)) %>%
  inner_join(ncbi_ortho_summary, by = c("ncbi_gene_id_num" = "salmon_gene_id"))

cat(sprintf("NCBI unnamed (LOC) genes with human ortholog: %d\n", nrow(ncbi_unnamed_ortho)))

ncbi_unnamed_ortho %>%
  count(biotype, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100) %>%
  head(10) %>%
  knitr::kable(
    caption = "Biotypes of unnamed NCBI genes with human orthologs",
    col.names = c("Biotype", "Count", "%"),
    digits = 1
  )
```

## Cross-annotation ortholog comparison

```{r cross_orthologs}
# Among matched gene pairs: ortholog in one source but not the other
ortho_cross <- genes_joined %>%
  filter(class %in% c("Green", "Yellow")) %>%
  count(ens_has_ortholog, ncbi_has_ortholog) %>%
  mutate(
    label = case_when(
      ens_has_ortholog & ncbi_has_ortholog ~ "Both have ortholog",
      ens_has_ortholog & !ncbi_has_ortholog ~ "Ensembl only",
      !ens_has_ortholog & ncbi_has_ortholog ~ "NCBI only",
      TRUE ~ "Neither"
    )
  )

ortho_cross %>%
  select(label, n) %>%
  knitr::kable(
    caption = "Human ortholog status for Green/Yellow gene pairs",
    col.names = c("Ortholog status", "Count")
  )
```


# 7. Ensembl Name Source Analysis

```{r desc_source_analysis}
# Breakdown of description sources for all Ensembl genes that have one
ens_with_desc <- ens_meta %>%
  filter(description != "NA" & !is.na(description))

desc_source_summary <- ens_with_desc %>%
  count(description_source, sort = TRUE) %>%
  mutate(pct = n / sum(n) * 100)

desc_source_summary %>%
  knitr::kable(
    caption = "Ensembl gene description sources",
    col.names = c("Source", "Count", "%"),
    digits = 1
  )
```

```{r source_vs_naming}
# Which sources correlate with having a Name= attribute?
source_naming <- ens_with_desc %>%
  group_by(description_source) %>%
  summarise(
    total = n(),
    named = sum(is_named),
    pct_named = named / total * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(total))

source_naming %>%
  knitr::kable(
    caption = "Description source vs having a Name attribute",
    col.names = c("Source", "Total", "Named", "% Named"),
    digits = 1
  )
```

```{r source_barplot, fig.height=5}
source_naming %>%
  filter(total >= 10) %>%
  mutate(description_source = fct_reorder(description_source, pct_named)) %>%
  ggplot(aes(x = description_source, y = pct_named, fill = total)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(trans = "log10", labels = comma) +
  labs(
    title = "Percentage of Genes with Name Attribute by Description Source",
    subtitle = "Bar color = total genes from that source (log scale)",
    x = "Description source",
    y = "% with Name attribute",
    fill = "Total genes"
  )
```

```{r hgnc_analysis}
# HGNC-sourced genes: are these derived from human orthologs?
hgnc_genes <- ens_meta %>%
  filter(description_source == "HGNC")

hgnc_with_ortho <- hgnc_genes %>%
  inner_join(ens_ortho_summary, by = c("gene_id" = "ensembl_gene_id"))

cat(sprintf("HGNC-sourced Ensembl genes: %d\n", nrow(hgnc_genes)))
cat(sprintf("  With human ortholog in BioMart: %d (%.1f%%)\n",
            nrow(hgnc_with_ortho), nrow(hgnc_with_ortho) / nrow(hgnc_genes) * 100))
```


# 8. Summary & Conclusions

```{r summary_numbers}
n_both_named <- sum(matched$ens_named & matched$ncbi_named, na.rm = TRUE)
n_ens_only_named <- sum(matched$ens_named & !matched$ncbi_named, na.rm = TRUE)
n_ncbi_only_named <- sum(!matched$ens_named & matched$ncbi_named, na.rm = TRUE)
n_neither_named <- sum(!matched$ens_named & !matched$ncbi_named, na.rm = TRUE)

concordance_rate <- if (nrow(both_named) > 0) {
  mean(both_named$names_match_ci, na.rm = TRUE) * 100
} else 0
```

## Key Findings

- **Total genes:** Ensembl has `r comma(ens_total)` gene-like features; NCBI has `r comma(ncbi_total)` genes
- **Named genes:** Ensembl `r comma(ens_named_total)` (`r round(ens_named_total/ens_total*100, 1)`%); NCBI `r comma(ncbi_named_total)` (`r round(ncbi_named_total/ncbi_total*100, 1)`%) with real symbols
- **Among matched gene pairs:**
  - Both named: `r comma(n_both_named)`
  - Ensembl only: `r comma(n_ens_only_named)`
  - NCBI only: `r comma(n_ncbi_only_named)`
  - Neither: `r comma(n_neither_named)`
- **Name concordance:** Where both are named, `r round(concordance_rate, 1)`% match (case-insensitive)
- **Human orthologs:** `r comma(n_ens_with_ortho)` Ensembl and `r comma(n_ncbi_with_ortho)` NCBI genes have human orthologs

## Main Reasons for Naming Asymmetry

1. **Ensembl names many non-coding genes** (via RFAM, miRBase) that NCBI only assigns LOC IDs
2. **NCBI names some protein-coding genes** that Ensembl annotates but doesn't assign a Name attribute to
3. **Biotype differences:** Ensembl annotates more lncRNA genes, which are rarely named in either annotation
4. **Source differences:** Ensembl draws names from ZFIN, HGNC, RFAM; NCBI uses its own nomenclature pipeline

## Recommendations for Salmobase

1. **Prefer Ensembl names** where available (broader coverage, standardized sources)
2. **Supplement with NCBI names** for genes named in NCBI but not Ensembl
3. **LOC genes with human orthologs** (`r comma(nrow(ncbi_unnamed_ortho))` in NCBI) are candidates for future naming via ortholog transfer
4. **Display both names** when they differ, so users can cross-reference

---

Report generated: `r Sys.time()`
